---
title: "RangeX_phenology_2023_data_exploration"
author: "Nadine Arzt"
format: html
editor: visual
date: 28.02.2024
---

## RangeX phenology 2023 data exploration

### Library

```{r}
library(tidyverse)

## set theme for poster with bigger font size
theme_set(theme_bw(base_size = 20))
```

### Import data set by sourcing the quarto file for cleaning

```{r}
#' @param file path to a quarto or Rmarkdown file
source_qmd <- function(file){
  tmp <- tempfile()
  knitr::purl(input = file, output = tmp)
  source(tmp)
}


source_qmd("RangeX_phenology_2023_data_cleaning.qmd")

```

### First plot

```{r}
ggplot(data = rangex_phenology_clean, aes(date, number_buds, color = site, group = unique_plant_id))+
  geom_point()+
  geom_line()+
  facet_wrap(vars(species))
```

### Subset of Plantago

```{r}
## filter only plalan
phenology_clean_plalan <- rangex_phenology_clean %>% 
  dplyr::filter(rangex_phenology_clean$species == "plalan")

## plot number of buds
ggplot(data = phenology_clean_plalan, aes(date, number_buds, color = site, group = unique_plant_id))+
  geom_point()+
  geom_line()
```

#### Color by treatment

```{r}
ggplot(data = phenology_clean_plalan, aes(date, number_buds, color = treatment, group = unique_plant_id))+
  geom_point()+
  geom_line()
```

#### Buds, flowers and infructescences

```{r}
ggplot(data = phenology_clean_plalan, aes(date)) +
  geom_point(aes(y = number_buds, color = "Number of Buds"), size = 3) +
  geom_point(aes(y = number_flowers, color = "Number of Flowers"), size = 3) +
  geom_point(aes(y = number_infructescences, color = "Number of Infructescences"), size = 3) +
  facet_wrap(~treatment) +
  labs(y = "Mean number of buds / flowers / infructescences", x = "Date") +
  scale_color_manual(values = c("lightblue", "darkred", "orange"),
                     name = "Variables",
                     labels = c("Number of buds", "Number of flowers", "Number of infructescences")) +
  theme(legend.position = "top")

## too many points in the plots
```

#### 

```{r}
ggplot(data = phenology_clean_plalan, aes(x = date)) +
  geom_point(aes(y = number_buds, color = "Number of Buds"), size = 3) +
  geom_errorbar(aes(ymin = number_buds - sd(number_buds), ymax = number_buds + sd(number_buds)), width = 0.1) +
  geom_point(aes(y = number_flowers, color = "Number of Flowers"), size = 3) +
  geom_errorbar(aes(ymin = number_flowers - sd(number_flowers), ymax = number_flowers + sd(number_flowers)), width = 0.1) +
  geom_point(aes(y = number_infructescences, color = "Number of Infructescences"), size = 3) +
  geom_errorbar(aes(ymin = number_infructescences - sd(number_infructescences), ymax = number_infructescences + sd(number_infructescences)), width = 0.1) +
  facet_wrap(~treatment) +
  labs(y = "Mean number of buds / flowers / infructescences", x = "Date") +
  scale_color_manual(values = c("lightblue", "darkred", "orange"),
                     name = "Variables",
                     labels = c("Number of buds", "Number of flowers", "Number of infructescences")) +
  theme(legend.position = "top")

```

#### Mean number of buds, flowers, infructescences

Calculate a mean per species per treatment per date and the variance as well.

```{r}
# Calculating mean for buds, flowers, infructescences
phenology_clean_plalan_mean_var <- phenology_clean_plalan %>%
  group_by(species, date, treatment) %>%
  summarise(
    mean_number_buds = mean(number_buds, na.rm = TRUE),
    mean_number_flowers = mean(number_flowers, na.rm = TRUE),
    mean_number_infructescences = mean(number_infructescences, na.rm = TRUE),
    sd_number_buds = sd(number_buds, na.rm = TRUE),
    sd_number_flowers = sd(number_flowers, na.rm = TRUE),
    sd_number_infructescences = sd(number_infructescences, na.rm = TRUE)
  ) %>%
  ungroup()

# Plotting
ggplot(phenology_clean_plalan_mean_var, aes(x = date)) +
  geom_pointrange(aes(y = mean_number_buds, ymin = mean_number_buds - sd_number_buds, ymax = mean_number_buds + sd_number_buds, color = "Number of Buds"), size = 1) +
  geom_pointrange(aes(y = mean_number_flowers, ymin = mean_number_flowers - sd_number_flowers, ymax = mean_number_flowers + sd_number_flowers, color = "Number of Flowers"), size = 1) +
  geom_pointrange(aes(y = mean_number_infructescences, ymin = mean_number_infructescences - sd_number_infructescences, ymax = mean_number_infructescences + sd_number_infructescences, color = "Number of Infructescences"), size = 1) +
  facet_wrap(~treatment) +
  labs(y = "Mean number of buds/flowers/infructescences", x = "") +
  scale_color_manual(values = c("darkblue", "darkred", "darkorange"),
                     name = "Variables",
                     labels = c("Number of Buds", "Number of Flowers", "Number of Infructescences")) +
  theme(legend.position = "top")

```

#### 

### Subset of Leucanthemum

```{r}
## filter only leuvul
phenology_clean_leuvul <- rangex_phenology_clean %>% 
  dplyr::filter(rangex_phenology_clean$species == "leuvul")

## plot number of buds
ggplot(data = phenology_clean_leuvul, aes(date, number_buds, color = site, group = unique_plant_id))+
  geom_point()+
  geom_line()
```

#### Mean number of buds, flowers, infructescences

```{r}
phenology_clean_leuvul_mean_var <- phenology_clean_leuvul %>%
  group_by(species, date, treatment) %>%
  summarise(
    mean_number_buds = mean(number_buds, na.rm = TRUE),
    mean_number_flowers = mean(number_flowers, na.rm = TRUE),
    mean_number_infructescences = mean(number_infructescences_adj, na.rm = TRUE),
    sd_number_buds = sd(number_buds, na.rm = TRUE),
    sd_number_flowers = sd(number_flowers, na.rm = TRUE),
    sd_number_infructescences = sd(number_infructescences_adj, na.rm = TRUE)
  ) %>%
  ungroup()

# Plotting
ggplot(phenology_clean_leuvul_mean_var, aes(x = date)) +
  geom_pointrange(aes(y = mean_number_buds, ymin = mean_number_buds - sd_number_buds, ymax = mean_number_buds + sd_number_buds, color = "Number of Buds"), size = 1, alpha = 0.7, position = position_dodge(width = 0.2)) +
  geom_pointrange(aes(y = mean_number_flowers, ymin = mean_number_flowers - sd_number_flowers, ymax = mean_number_flowers + sd_number_flowers, color = "Number of Flowers"), size = 1, alpha = 0.7, position = position_dodge(width = 0.5)) +
  geom_pointrange(aes(y = mean_number_infructescences, ymin = mean_number_infructescences - sd_number_infructescences, ymax = mean_number_infructescences + sd_number_infructescences, color = "Number of Infructescences"), size = 1, alpha = 0.7, position = position_dodge(width = 0.8)) +
  facet_wrap(~treatment) +
  labs(y = "Mean number of buds/flowers/cum infructescences", x = "") +
  scale_color_manual(values = c("darkblue", "darkred", "darkorange"),
                     name = "Variables",
                     labels = c("Number of Buds", "Number of Flowers", "Number of Infructescences")) +
  theme(legend.position = "top")
```

### Subset of Trifolium

```{r}
## filter only tripra
phenology_clean_tripra <- rangex_phenology_clean %>% 
  dplyr::filter(rangex_phenology_clean$species == "tripra")

## plot number of buds
ggplot(data = phenology_clean_tripra, aes(date, number_buds, color = site, group = unique_plant_id))+
  geom_point()+
  geom_line()
```

#### Mean number of buds, flowers, infructescences

```{r}
phenology_clean_tripra_mean_var <- phenology_clean_tripra %>%
  group_by(species, date, treatment) %>%
  summarise(
    mean_number_buds = mean(number_buds, na.rm = TRUE),
    mean_number_flowers = mean(number_flowers, na.rm = TRUE),
    mean_number_infructescences = mean(number_infructescences_adj, na.rm = TRUE),
    sd_number_buds = sd(number_buds, na.rm = TRUE),
    sd_number_flowers = sd(number_flowers, na.rm = TRUE),
    sd_number_infructescences = sd(number_infructescences_adj, na.rm = TRUE)
  ) %>%
  ungroup()

# Plotting
ggplot(phenology_clean_tripra_mean_var, aes(x = date)) +
  geom_pointrange(aes(y = mean_number_buds, ymin = mean_number_buds - sd_number_buds, ymax = mean_number_buds + sd_number_buds, color = "Number of Buds"), size = 1, alpha = 0.7, position = position_dodge(width = 0.2)) +
  geom_pointrange(aes(y = mean_number_flowers, ymin = mean_number_flowers - sd_number_flowers, ymax = mean_number_flowers + sd_number_flowers, color = "Number of Flowers"), size = 1, alpha = 0.7, position = position_dodge(width = 0.5)) +
  geom_pointrange(aes(y = mean_number_infructescences, ymin = mean_number_infructescences - sd_number_infructescences, ymax = mean_number_infructescences + sd_number_infructescences, color = "Number of Infructescences"), size = 1, alpha = 0.7, position = position_dodge(width = 0.8)) +
  facet_wrap(~treatment) +
  labs(y = "Mean number of buds/flowers/cum infructescences", x = "") +
  scale_color_manual(values = c("darkblue", "darkred", "darkorange"),
                     name = "Variables",
                     labels = c("Number of Buds", "Number of Flowers", "Number of Infructescences")) +
  theme(legend.position = "top")
```

#### Plot the range instead of points

```{r}
ggplot(phenology_clean_tripra_mean_var, aes(x = date)) +
  geom_line(aes(y = mean_number_buds, color = "Number of Buds"))  +
  geom_ribbon(aes(ymin = mean_number_buds - sd_number_buds, ymax = mean_number_buds + sd_number_buds, fill = "blue"))+
  facet_wrap(~treatment) +
  labs(y = "Mean number of buds/flowers/cum infructescences", x = "") +
  scale_color_manual(values = c("darkblue", "darkred", "darkorange"),
                     name = "Variables",
                     labels = c("Number of Buds", "Number of Flowers", "Number of Infructescences")) +
  theme(legend.position = "top")

## problem: it plots negative values, but in the table are no negative values
```

### Subset of Succisa

```{r}
## filter only sucpra
phenology_clean_sucpra <- rangex_phenology_clean %>% 
  dplyr::filter(rangex_phenology_clean$species == "sucpra")

## plot number of buds
ggplot(data = phenology_clean_sucpra, aes(date, number_buds, color = site, group = unique_plant_id))+
  geom_point()+
  geom_line()
```

#### Mean number of buds, flowers, infructescences

```{r}
phenology_clean_sucpra_mean_var <- phenology_clean_sucpra %>%
  group_by(species, date, treatment) %>%
  summarise(
    mean_number_buds = mean(number_buds, na.rm = TRUE),
    mean_number_flowers = mean(number_flowers, na.rm = TRUE),
    mean_number_infructescences = mean(number_infructescences_adj, na.rm = TRUE),
    sd_number_buds = sd(number_buds, na.rm = TRUE),
    sd_number_flowers = sd(number_flowers, na.rm = TRUE),
    sd_number_infructescences = sd(number_infructescences_adj, na.rm = TRUE)
  ) %>%
  ungroup()

# Plotting
ggplot(phenology_clean_sucpra_mean_var, aes(x = date)) +
  geom_pointrange(aes(y = mean_number_buds, ymin = mean_number_buds - sd_number_buds, ymax = mean_number_buds + sd_number_buds, color = "Number of Buds"), size = 1, alpha = 0.7, position = position_dodge(width = 0.2)) +
  geom_pointrange(aes(y = mean_number_flowers, ymin = mean_number_flowers - sd_number_flowers, ymax = mean_number_flowers + sd_number_flowers, color = "Number of Flowers"), size = 1, alpha = 0.7, position = position_dodge(width = 0.5)) +
  geom_pointrange(aes(y = mean_number_infructescences, ymin = mean_number_infructescences - sd_number_infructescences, ymax = mean_number_infructescences + sd_number_infructescences, color = "Number of Infructescences"), size = 1, alpha = 0.7, position = position_dodge(width = 0.8)) +
  facet_wrap(~treatment) +
  labs(y = "Mean number of buds/flowers/cum infructescences", x = "") +
  scale_color_manual(values = c("darkblue", "darkred", "darkorange"),
                     name = "Variables",
                     labels = c("Number of Buds", "Number of Flowers", "Number of Infructescences")) +
  theme(legend.position = "top")
```

## Pivot longer the data

```{r}
rangex_phenology_clean_long <- rangex_phenology_clean |> 
  pivot_longer(cols = c(number_buds, number_flowers, number_infructescences_adj),
               names_to = "stage",
               values_to = "value")

# View the resulting dataframe
print(rangex_phenology_clean_long)
```

#### Filter Sucpra

```{r}
## filter only sucpra
phenology_clean_long_sucpra <- rangex_phenology_clean_long %>% 
  dplyr::filter(rangex_phenology_clean_long$species == "sucpra")

```

#### Calculate median and quantiles

```{r}
phenology_clean_sucpra_median_quant <- phenology_clean_long_sucpra |> 
  group_by(species, date, treatment, stage) |> 
  summarise(median = median(value),
    lo = quantile(value, probs = 0.1, na.rm = TRUE), 
    hi = quantile(value, probs = 0.9, na.rm = TRUE), 
    .groups = "drop") 
```

#### Plot range with median, min, max

```{r}
ggplot(phenology_clean_sucpra_median_quant, aes(x = date)) +
  geom_line(aes(y = median, color = stage)) +
  geom_ribbon(aes(y = median, ymin = lo, ymax = hi, fill = stage), size = 1, alpha = 0.5) +
  facet_wrap(~treatment) +
  labs(y = "Median", x = "") +
  scale_color_manual(values = c("darkblue", "darkred", "darkorange")) +
  scale_fill_manual(values = c("darkblue", "darkred", "darkorange")) +
  theme(legend.position = "top")

```

#### 
