---
title: "RangeX_phenology_2023_#buds_flowers_infructescences"
author: "Nadine Arzt"
format: html
editor: visual
---

### 

## RangeX phenology 2023 plot buds, flowers, infructescences

### Library

```{r}
library(tidyverse)
library(conflicted)
library(RColorBrewer)
library(viridis)

conflict_prefer_all("dplyr", quiet = TRUE)

## set theme for poster with bigger font size
theme_set(theme_bw(base_size = 20))
```

### Import data set by sourcing the quarto file for cleaning

```{r}
#' @param file path to a quarto or Rmarkdown file
source_qmd <- function(file){
  tmp <- tempfile()
  knitr::purl(input = file, output = tmp)
  source(tmp)
}


source_qmd("RangeX_phenology_2023_data_cleaning.qmd")
```

### Filter buds, flowers, infructescences

```{r}
## don't use seeds_collected
phenology_clean <- rangex_phenology_clean_long |> 
    filter(rangex_phenology_clean_long$stage != "seeds_collected")
```

### Calculate median and quantiles

```{r}
## calculate median of number of buds, flowers, infructescences
## calculate quantiles (fits to median) instead of sd to avoid having negative values for mean-/+ sd. 
## Negative ranges come when we have many low values and a few high ones
phenology_clean_median_quant <- phenology_clean |> 
  group_by(species, date, treatment, stage) |> 
  summarise(median = median(value),
    lo = quantile(value, probs = 0.1, na.rm = TRUE), 
    hi = quantile(value, probs = 0.9, na.rm = TRUE), 
    .groups = "drop")
```

### Rename treatment and stages

```{r}
## call number_buds buds ...
phenology_clean_median_quant <- phenology_clean_median_quant |> 
  mutate(stage = case_when(
    stage == "number_buds" ~ "buds",
    stage == "number_flowers" ~ "flowers",
    stage == "number_infructescences" ~ "infructescences",
    TRUE ~ as.character(stage)  # Keep other values as they are
  ))

## split column treatment
phenology_clean_median_quant <- phenology_clean_median_quant |> 
  separate(treatment, c("site", "temperature", "competition"))

## rename site
phenology_clean_median_quant <- phenology_clean_median_quant |> 
  mutate(site = case_when(
    site == "lo" ~ "low",
    site == "hi" ~ "high",
    TRUE ~ as.character(stage)  # Keep other values as they are
  ))

## rename vege = with competition
## rename bare = without competition
phenology_clean_median_quant <- phenology_clean_median_quant |> 
  mutate(competition = case_when(
    competition == "vege" ~ "with competition",
    competition == "bare" ~ "without competition",
    TRUE ~ as.character(stage)  # Keep other values as they are
  ))

## rename temperature
phenology_clean_median_quant <- phenology_clean_median_quant |> 
  mutate(temperature = case_when(
    temperature == "ambi" ~ "ambient",
    temperature == "warm" ~ "warmed",
    TRUE ~ as.character(stage)  # Keep other values as they are
  ))

## combine temperature and competition
phenology_clean_median_quant <- phenology_clean_median_quant |> 
  unite(site_treatment, site, temperature, sep = " ", remove = FALSE)

```

### First plot with flowers of all species

```{r}
## filter only flowers
phenology_clean_median_quant_flowers <- phenology_clean_median_quant |> 
  filter(phenology_clean_median_quant$stage == "flowers")

## filter only high site
# phenology_clean_median_quant_flowers_hi <- phenology_clean_median_quant_flowers |> 
#   filter(phenology_clean_median_quant_flowers$site == "high")

## combine temperature and competition
phenology_clean_median_quant_flowers <- phenology_clean_median_quant_flowers |> 
  unite(treatment, site_treatment, competition, sep = " ", remove = FALSE)


ggplot(phenology_clean_median_quant_flowers, aes(x = date)) +
  geom_line(aes(y = median, color = treatment)) +
  geom_ribbon(aes(y = median, ymin = lo, ymax = hi, fill = treatment), linewidth = 2, alpha = 0.4) +
  facet_wrap(vars(species)) +  # Added `scales = "free"`
  labs(y = "Median +/- quantiles", x = "", title = "") +
  scale_color_manual(values = c("darkblue", "darkred", "darkorange", "grey", "green", "yellow")) +
  scale_fill_manual(values = c("darkblue", "darkred", "darkorange", "grey", "green", "yellow")) +
  theme(legend.position = "top")
```

Problem with cynci hi 7a f9 –\> was g9 in metadata file –\> needs to be changed in google docs

```{r}
flowers_colors <- brewer.pal(6, "Set3")

## plot median of flower number for all species and all treatments
ggplot(data = phenology_clean_median_quant_flowers, aes(date, color = treatment))+
  geom_point(aes(y = median, color = treatment), size = 3)+
  geom_line(aes(y = median, color = treatment))+
  facet_wrap(vars(species), nrow = 2, ncol = 5)+
  labs(y = "Median", x = "", title = "") +
  theme(legend.position = "top")+
  scale_color_manual(values = flowers_colors)
```

### Filter per species

#### Subset of Plantago

```{r}
## filter only plalan
phenology_clean_median_quant_plalan <- phenology_clean_median_quant |> 
  filter(phenology_clean_median_quant$species == "plalan")

```

##### Plot

```{r}
# Convert site_treatment to a factor with correct order
phenology_clean_median_quant_plalan$site_treatment <- factor(
  phenology_clean_median_quant_plalan$site_treatment,
  levels = c("high warmed", "high ambient", "low ambient")
)

ggplot(phenology_clean_median_quant_plalan, aes(x = date)) +
  geom_line(aes(y = median, color = stage)) +
  geom_ribbon(aes(y = median, ymin = lo, ymax = hi, fill = stage), linewidth = 2, alpha = 0.4) +
  facet_grid(rows = vars(site_treatment), cols = vars(competition)) +  # Added `scales = "free"`
  labs(y = "Median +/- quantiles", x = "", title = "Plantago lanceolata") +
  scale_color_manual(values = c("darkblue", "darkred", "darkorange")) +
  scale_fill_manual(values = c("darkblue", "darkred", "darkorange")) +
  theme(legend.position = "top")
```

##### Play with different colors

```{r}
ggplot(phenology_clean_median_quant_plalan, aes(x = date)) +
  geom_line(aes(y = median, color = stage)) +
  geom_ribbon(aes(y = median, ymin = lo, ymax = hi, fill = stage), linewidth = 2, alpha = 0.5) +
  facet_grid(rows = vars(site_treatment), cols = vars(competition)) +  # Added `scales = "free"`
  labs(y = "Median +/- quantiles", x = "", title = "Plantago lanceolata") +
  scale_color_manual(values = c("turquoise3", "hotpink3", "orange1")) +
  scale_fill_manual(values = c("turquoise3", "hotpink3", "orange1")) +
  theme(legend.position = "top")
```
